Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-03 ind_files/04_unification_discussion.tex
Minor phrasing changes

where we have defined the statistical variance of a quantity $f$ as
\sigma^2_{f} \coloneqq \overline{(f^2)} - (\overline{f})^2

@@

Stochastic finite population models often exhibit behaviors that are markedly different from their deterministic limits. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of eco-evolutionary theory, \cite{doebeli_towards_2017} have called for a reformulation of evolutionary dynamics starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic, a view that has also been echoed by theorists before \citep{metcalf_why_2007,geritz_mathematical_2012}. In this thesis, I present and derive from first principles some fundamental relations (equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance}) that any such reformulation must satisfy. Further, these relations deal with biologically important quantities, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of \cite{queller_fundamental_2017}), or `unifying principles' (in the sense of \cite{lion_theoretical_2018}) for the dynamics of finite populations. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). Just like in infinite population models, the various terms of equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} lend themselves to simple biological interpretation. Note that actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that high dimensional evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos \citep{doebeli_diversity_2017}. We have also neglected any potential complications introduced by genotype-phenotype maps and other such genetic factors, since they can be absorbed into the per-capita birth and death rates. Nevertheless, our equations will still hold for arbitrarily complicated birth and death rates. Like the Price equation, the utility of these equations lies not in their solutions, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations \citep{frank_natural_2012, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature  \citep{queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision \citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede specific models of specific systems. In our case, our framework complements the various specific studies of social evolution in specific models \citep{constable_demographic_2016, houchmandzadeh_selection_2012, chotibut_evolutionary_2015, mcleod_social_2019} that pointed to noise-induced selection reversing the direction of evolution predicted by deterministic models.\\

@@

At first glance, the idea that individuals with lower birth and death rates are preferred over individuals with higher birth and death rates may be reminiscent of ideas like $r$ vs $K$ selection or selection for the pace of life in life-history evolution \citep{stearns_evolution_1977}. However, it is unclear whether this similarity reflects some deep principle or whether it is just superficial. Models in life-history theory are often primarily concerned with spatiotemporally fluctuating (external) environments, and thus the stochasticity in those models is extrinsic to the population. I have entirely neglected such extrinsic factors in my formalisms. In principle, it is possible to make the birth and death rates \ref{nD_functional_forms_for_replicator} in our model also depend on a temporally varying external environment $E(t)$ (whose variation may possibly depend on the population $\mathbf{v}(t)$). Incorporating such a term would ensure that the `ecological feedback' terms in equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} are non-zero, but may also lead to much more complex dynamics. Furthermore, if the environment $E(t)$ has some associated stochasticity, the final dynamics would be the result of interactions between two qualitatively different forms of noise --- \emph{extrinsic} noise from the environment, and \emph{intrinsic} noise from the finiteness of the population --- and thus will be rather complicated and likely analytically intractable. Ecological frameworks such as modern coexistence theory \citep{chesson_multispecies_1994}, which deal with questions about ecological population dynamics and would benefit from a first principles stochastic birth-death formulation, also generally work with fluctuating external environments. Thus, while the idea of integrating my birth-death framework with ecological ideas such as the pace-of-life syndrome \citep{mathot_models_2018, wright_life-history_2019} or modern coexistence theory \citep{chesson_multispecies_1994, johnson_resolving_2022} is biologically appealing, it is far from trivial and may present a promising avenue for future work.

>>>>

@@

Stochastic finite population models often exhibit behaviors that are markedly different from their deterministic limits. Since real-life populations are stochastic and finite, it is thus imperative that modellers work with stochastic first-principles models instead of their deterministic limits, lest they risk missing important phenomena that are unique to stochastic systems \citep{black_stochastic_2012,schreiber_does_2022,hastings_transients_2004,shoemaker_integrating_2020}. In the context of eco-evolutionary theory, \cite{doebeli_towards_2017} have called for a reformulation of evolutionary dynamics starting from stochastic birth-death processes on the grounds that such a formulation is more fundamental and mechanistic, a view that has also been echoed by theorists before \citep{metcalf_why_2007,geritz_mathematical_2012}. In this thesis, I present and derive from first principles some fundamental relations (equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance}) that any such reformulation must satisfy. Further, these relations deal with biologically important quantities, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of \cite{queller_fundamental_2017}), or `unifying principles' (in the sense of \cite{lion_theoretical_2018}) for the dynamics of finite populations. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - Equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). Just like in infinite population models, the various terms of equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} lend themselves to simple biological interpretation. Note that actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that high dimensional evolutionary birth-death models can exhibit a dizzying array of complicated phenomena, including limit cycles and evolutionary chaos \citep{doebeli_diversity_2017}. We have also neglected any potential complications introduced by genotype-phenotype maps and other such genetic factors, since they can be absorbed into the per-capita birth and death rates. Nevertheless, our equations will still hold for arbitrarily complicated birth and death rates. Like the Price equation, the utility of these equations lies not in their solutions, but instead in their generality and the fact that their terms help us clearly understand the various forces operating in biological populations \citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general approach of working with `high-level' processes without specifying system-specific details is part of a broader pursuit of `model-independent' eco-evolutionary theory that has recently been gaining popularity in the literature  \citep{grafen_formal_2014, queller_fundamental_2017, doebeli_towards_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022}. Such model-independent descriptions that `abstract away' system-specific details almost inevitably come at the cost of precision \citep{levins_strategy_1966, potochnik_idealization_2018}, and are thus intended to complement rather than supersede specific models of specific systems. In our case, our framework complements the various specific studies of social evolution in specific models \citep{constable_demographic_2016, houchmandzadeh_selection_2012, chotibut_evolutionary_2015, mcleod_social_2019} that pointed to noise-induced selection reversing the direction of evolution predicted by deterministic models.\\

@@

At first glance, the idea that individuals with lower birth and death rates are preferred over individuals with higher birth and death rates may be reminiscent of ideas in life-history evolution like $r$ vs $K$ selection or selection for the pace of life in stochastic models  \citep{stearns_evolution_1977}. However, it is unclear whether this similarity reflects some deep principle or whether it is just superficial. Models in life-history theory are often primarily concerned with spatiotemporally fluctuating (external) environments, and thus the stochasticity in those models is extrinsic to the population. I have entirely neglected such extrinsic factors in my formalisms. In principle, it is possible to make the birth and death rates \ref{nD_functional_forms_for_replicator} in our model also depend on a temporally varying external environment $E(t)$ (whose variation may possibly depend on the population $\mathbf{v}(t)$). Incorporating such a term would ensure that the `ecological feedback' terms in equations \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} are non-zero, but may also lead to much more complex dynamics. Furthermore, if the variation of the environment $E(t)$ has some associated stochasticity, the final dynamics would be the result of interactions between two qualitatively different forms of noise --- \emph{extrinsic} noise from the environment, and \emph{intrinsic} noise from the finiteness of the population --- and thus will be rather complicated and likely analytically intractable. Ecological frameworks such as modern coexistence theory \citep{chesson_multispecies_1994}, which deal with questions about ecological population dynamics and would benefit from a first principles stochastic birth-death formulation, also generally work with fluctuating external environments. Thus, while the idea of integrating my birth-death framework with ecological ideas such as the pace-of-life syndrome \citep{mathot_models_2018, wright_life-history_2019} or modern coexistence theory \citep{chesson_multispecies_1994, johnson_resolving_2022} is biologically appealing, it is far from trivial and may present a promising avenue for future work.
